Paper template for {ICPhS} 2023 Prague

### 2024-04-27 manuscript/main.tex

Write something here

@@

Both Catalan and Peninsular Spanish sound inventories have an phoneme
/s/, which can be classified as a voiceless alveolar fricative
\cite{hualde_colina_2015}. In addition, most of the Catalan varieties
have a voiced alveolar phoneme in intervocalic contexts \textbf{(Lloret,
2011; Recasens Vives, 1996, p.~279)}. In onset position, /s/ and /z/
contrast phonemically, which leads to minimal pairs like \emph{cel}
{[}s\textepsilon l{]} - \emph{zel} {[}z\textepsilon l{]} and
\emph{fossa} {[}fósa{]} - \emph{fosa} {[}fóza{]} \textbf{(Carrera-Sabaté
\& Prieto, 2012-2019)}. Conversely, at the end of the word, this
contrast is neutralized, being the voicing or devoicing of the alveolar
fricative dependent on the following sound \textbf{(Institut d'Estudis
Catalans, 2022)}. For example, if it is followed by a vowel, the
alveolar fricative is voiced: \emph{torres} {[}tór\textschwa{}s{]} and
\emph{torres antigues}
{[}tor\textschwa{}z\textschwa{}nti\textgamma{}\textschwa{}s{]}.

@@

There is some evidence of the influence of Spanish on the production of
the Catalan alveolar fricative, in contexts where, according to
prescriptivist rules \textbf{(Institut d'Estudis Catalans, 2022)}, the
{[}z{]} should be produced, but {[}s{]} is produced instead. This
phenomenon has mainly been attested in those areas where there has been
an extended contact between Catalan and Spanish, such asthe so-called
``franja de Aragón'' (borderland between Catalonia and Aragon), some
areas of the Valencian Community, and Barcelona and other urban areas
\textbf{(Benet et al., 2012; Davidson, 2020; Recasens Vives, 1996,
p.~268)}. The opposite pattern has also been observed - that is, the
voicing of the alveolar fricative in the Spanish spoken in Barcelona and
Valencia. Interestingly, although in both Barcelona and Valencia Catalan
and Spanish coexist and influence each other, it seems that Catalan has
a greater influence on Spanish in Barcelona, but Spanish has a greater
influence on Catalan in Valencia \textbf{(Davidson, 2020)}.

@@

Participant responses to the statements \emph{Parla bé català} `S/he
speaks proper Catalan' and \emph{És d'origen català} `S/he is of Catalan
origin' showed similar results, as illustrated in Figures \ref{fig:pbc}
and \{fig:oc\}. Those who have both Catalan and Spanish as their L1
rated the speaker more positively regardless of phoneme, compared to
those whose L1 is either Catalan or Spanish.

@@

The aim of this exploration study was to investigate whether phoneme
production ({[}s{]} vs {[}z{]}), mother tongue (Catalan vs Spanish vs
both), and province of origin (Girona vs Barcelona) influenced the
attitudes of the participants towards a Catalan native speaker. The
results suggest that these predictors have a different effect on the
ratings of the speaker depending on the social characteristic under
evaluation.

Contrary to what we expected, phoneme was not a significant predictor.
One explanation for this divergent pattern could be that either
participants could not distinguish between {[}s{]} and {[}z{]} or, if
they could, this difference in production was not changing their
attitudes towards the speaker. This finding is interesting considering
that, in our stimuli, the production of the voiceless alveolar fricative
in intervocalic position has been considered a result of Catalan in
contact with Spanish \textbf{(Benet et al., 2012; Davidson, 2020;
Recasens Vives, 1996, p.~268)}, and, from a prescriptivist point of
view, it should be avoided \textbf{(Institut d'Estudis Catalans, 2022)}.
Despite the lack of significance of this predictor, it should be
mentioned that the speaker received more positive ratings when the
{[}z{]} was being produced. Although at this point we can only
hypothesize, it could be that the combination of the phoneme /s/ with
the precedent vowel, might give the listener more cues and have a
stronger social meaning. Future studies should address this issue in
order to explore if the combination of vowel + /s/ carries other social
connotations. \textbf{REVISAR}

>>>>

Catalan and Spanish are two languages that have coexisted for centuries,
mainly in eastern Spain, along the Mediterranean cost \textbf{(Hualde,
1992)}. It is safe to assume that there is a bidirectional influence
between these languages: in that region, Catalan has influenced Spanish
and Spanish has influenced Catalan in all linguistic levels
\textbf{(Davidson, 2014, 2015)}. A crucial area of cross-linguistic
influence might be the production of sounds, especially in those sounds
who have different distribution in Catalan and Spanish. For example,
previous studies have found that, in Barcelona, the production of
Catalan vowels is different based on the L1 of the speaker, being it
less target-like in L1 Spanish speakers compared to L1 Catalan speakers
\textbf{(CITAR SEBASTIAN GALLES, etc)}.

As for how Catalan people perceive these differences in production and,
thus, their attitudes towards a less Catalan-like production of a given
phoneme little is known. The aim of this study is to examine the covert
attitudes towards the production of intervocalic /s/, as either {[}s{]}
(disfavored by prescriptivist norms) or {[}z{]} (preferred by
prescriptivist norms) by L1 Catalan, L1 Spanish, and Catalan-Spanish
bilinguals from Catalonia using a matched guise test \textbf{(Lambert,
1967)}. The alveolar fricative was chosen because it exists in both
Catalan and Spanish inventories, but their contextual distribution is
different (Llisterri, 2023).

@@

Both Catalan and Peninsular Spanish sound inventories have a phoneme
/s/, which can be classified as a voiceless alveolar fricative
\cite{hualde_colina_2015}. In addition, most of the Catalan varieties
have a voiced alveolar phoneme in intervocalic contexts \textbf{(Lloret,
2011; Recasens Vives, 1996, p.~279)}. In onset position, /s/ and /z/
contrast phonemically, which leads to minimal pairs like \emph{cel}
{[}s\textepsilon l{]} - \emph{zel} {[}z\textepsilon l{]} and
\emph{fossa} {[}fósa{]} - \emph{fosa} {[}fóza{]} \textbf{(Carrera-Sabaté
\& Prieto, 2012-2019)}. Conversely, at the end of the word, this
contrast is neutralized, being the voicing or devoicing of the alveolar
fricative dependent on the following sound \textbf{(Institut d'Estudis
Catalans, 2022)}. For example, if it is followed by a vowel, the
alveolar fricative is voiced: \emph{torres} {[}tór\textschwa{}s{]} and
\emph{torres antigues}
{[}tor\textschwa{}z\textschwa{}nti\textgamma{}\textschwa{}s{]}.

@@

There is some evidence of the influence of Spanish on the production of
the Catalan alveolar fricative, in contexts where, according to
prescriptivist rules \textbf{(Institut d'Estudis Catalans, 2022)}, the
{[}z{]} should be produced, but {[}s{]} is produced instead. This
phenomenon has mainly been attested in those areas where there has been
an extended contact between Catalan and Spanish, such as the so-called
``franja de Aragón'' (borderland between Catalonia and Aragon), some
areas of the Valencian Community, and Barcelona and other urban areas
\textbf{(Benet et al., 2012; Davidson, 2020; Recasens Vives, 1996,
p.~268)}. The opposite pattern has also been observed - that is, the
voicing of the alveolar fricative in the Spanish spoken in Barcelona and
Valencia. Interestingly, although in both Barcelona and Valencia Catalan
and Spanish coexist and influence each other, it seems that Catalan has
a greater influence on Spanish in Barcelona, but Spanish has a greater
influence on Catalan in Valencia \textbf{(Davidson, 2020)}.

@@

Participant responses to the statements \emph{Parla bé català} `S/he
speaks proper Catalan' and \emph{És d'origen català} `S/he is of Catalan
origin' show similar results, as illustrated in Figures \ref{fig:pbc}
and \{fig:oc\}. Those who have both Catalan and Spanish as their L1 rate
the speaker more positively regardless of phoneme, compared to those
whose L1 is either Catalan or Spanish. In addition, although it is not
statistically significant, participants give higher ratings to the
speaker when they are listening to those statements with {[}z{]}.

@@

The aim of this exploration study was to investigate whether phoneme
production ({[}s{]} vs {[}z{]}), mother tongue (Catalan vs Spanish vs
both), and province of origin (Girona vs Barcelona) influenced the
attitudes of the participants towards a native speaker of Central
Catalan. The results suggest that these predictors have a different
effect on the ratings of the speaker depending on the social
characteristic under evaluation.

Contrary to what we expected, phoneme was not a significant predictor.
One explanation for this divergent pattern could be that either
participants could not distinguish between {[}s{]} and {[}z{]}
\textbf{(Davidson, 2019)} or, if they could, this difference in
production was not changing their attitudes towards the speaker
\textbf{(Sinner, 2002)}. This finding is interesting considering that
the production of the voiceless alveolar fricative in intervocalic
position has been considered a result of Catalan in contact with Spanish
\textbf{(Benet et al., 2012; Davidson, 2020; Recasens Vives, 1996,
p.~268)}, and, from a prescriptivist point of view, it should be avoided
\textbf{(Institut d'Estudis Catalans, 2022)}. Despite the lack of
significance of this predictor, it should be mentioned that the speaker
received more positive ratings when the {[}z{]} was being produced. The
fact that phoneme was not an important factor contrasts with
\textbf{Davidson (2020)}, who analyzed covert and overt attitudes
towards Catalonian Spanish and found that {[}z{]} was a significant
predictor: it was considered a sociolinguistic indicator of
Catalan-Spanish bilingual membership. Although at this point we can only
hypothesize, it could be that {[}z{]} is perceptually more salient in
Spanish than in Catalan, as the Spanish inventory does not have the
phoneme /z/, but voicing of /s/ might partially occur \textbf{(Benet et
al., 2012; Hualde \& Colina, 2015; Hualde et al., 2020)}. Another
explanation could be that the /s/ alone does not provide enough social
information; rather, the combination of the phoneme /s/ with the
precedent vowel might give the listener more cues and have a stronger
social meaning. Future studies should address this issue in order to
explore if the combination of vowel + /s/ carries other social
connotations than just the /s/ alone.

We expected mother tongue to be a strong predictor of the attitudes
towards the production of intervocalic /s/, specifically, that L1
Catalan and bilingual speakers would give higher rates to those
statements with {[}z{]}. This hypothesis was only partially confirmed.
As reported above, phoneme was not a factor influencing the attitudes,
and mother tongue was only a significant predictor in the two statements
related to Catalan identity, namely \emph{Parla bé català} `S/he speaks
proper Catalan' and \emph{És d'origen català} `S/he is of Catalan
origin'. Interestingly, bilinguals gave higher ratings to the speaker,
followed by L1 Catalan participants and L1 Spanish participants. A
potential explanation for this finding, that should be verified in
future research, could be that L1 Catalan speakers have specific ideas
(higher standards) towards the `proper' Catalan. In addition, some
studies have suggested that those Catalan people who grow up in
Spanish-speaking households typically exhibit more negative attitudes
towards Catalan \textbf{(CITATIONS)}, which is consistent with our
results.

Lastly, province was also found as an important predictor, but only for
some statements. In two of the solidarity statements (\emph{És simpàtic}
`S/he is nice'; \emph{És de fiar} `S/he is trustworthy'), the speaker
was rated more positively by those participants from the province of
Barcelona. This finding might suggest that they identify with the
speaker; thus, giving her higher ratings \textbf{BUSCAR CITA}.
Interestingly, the speaker that created the stimuli is from the province
of Girona, but has spent the last 11 years of her life living outside
that area, either in Barcelona or in another country. We asked her to
read the sentences using a Central Catalan `neutral accent', which
people might have potentially related with Barcelona. This `neutral
accent' does not show any of the prototypical characteristics of the
\emph{accent gironí}, a subvariety within Central Catalan spoken in some
areas of Girona \textbf{(Gros, 2024)}.

In summary, our findings suggest that the relationship between covert
attitudes and speaker ratings is not straightforward, as the speaker was
evaluated differently based on the social characteristic, with varying
factors influencing such ratings (mother tongue and province).
Crucially, phoneme production did not had an effect on the ratings, even
though prescriptivist norms disfavor the realization of {[}s{]} in
intervocalic position. Future studies should include a larger
participant pool, as well as participants from other regions in
Catalonia, such as Lleida and Tarragona. In addition, examining other
phoneme productions that emerge through the contact of Catalan with
Spanish could give us a better understanding of the factors that affect
the attitudes towards the Catalan language, and potential
sociolinguistic stereotypes and indicators related to Catalan identity.
